Fluent statistical computing interfaces for biological data analysis

### 2020-07-26 Rmd/02-plyranges.Rmd
update toc; rebuild
update text
fix spelling
consistency of formatting

There is a cognitive load placed on users in learning a data abstraction from the Bioconductor project and understanding its appropriate use. Users must navigate these abstractions to perform a genomic analysis task, when a single data abstraction, a GRanges object will suffice. By recognizing that the GRanges class follows 'tidy' data principles, we create a grammar of genomic data transformation, defining verbs for performing actions on and between genomic interval data and providing a way of performing common data analysis tasks through a coherent interface to existing Bioconductor infrastructure. We implement this grammar as a Bioconductor/R package called `plyranges`.

@@

Several domain specific languages (DSLs) enable scientists to process and reason
about heterogeneous genomics data by expressing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED. An example of the former is the Genome Query 
Language (GQL) and its distributed implementation GenAp which use a 
SQL-like syntax for fast retrieval of information of unprocessed 
sequencing data [@Kozanitis2014-va; @Kozanitis2016-bm]. Similarly, the Genometric 
Query Language (GMQL) implements a DSL for combining genomic datasets 
[@Kaitoua2017-pw].  The command line application BEDtools develops an extensive 
algebra for performing arithmetic between two or more sets of genomic regions 
[@Quinlan2010-gc]. All of the aforementioned DSLs are designed to be evaluated
either at the command line or embedded in scripts for batch processing. They 
exist in a sparse ecosystem, mostly consisting of UNIX and database tools that
lack biological semantics and operate at the level of files and
database tables.

The Bioconductor/R packages `IRanges` and `GenomicRanges` 
[@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
analyzing genomics data with R, an interactive data analysis environment that 
encourages reproducibility and provides high-level abstractions for manipulating,
modelling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs 
that are embedded in programming languages include pybedtools and valr 
[@Dale2011-js; @Kent2017], however these packages lack the interoperability
provided by the aforementioned Bioconductor packages and are not easily 
extended.

@@

Recently, there has been a concerted, community effort to standardize
R data structures and workflows around the notion of tidy data
[@Wickham2014-jc]. A tidy dataset is defined as a tabular
data structure that has observations as rows and columns as variables,
and all measurements pertain to a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relate to the design of an experiment
and the variables measured. The `dplyr` package [@Wickham2017-dplyr] 
defines an application programming interface (API) that maps notions from 
the general relational algebra to verbs that act on tidy data. These verbs
can be composed together on one or more tidy datasets with the pipe operator 
from the `magrittr` package [@R-magrittr]. Taken together these features 
enable a user to write human readable analysis workflows.

We have created a genomic DSL called `plyranges` that reformulates
notions from existing genomic algebras and embeds them in R as a
genomic extension of `dplyr`. By analogy, `plyranges` is to the
genomic algebra, as `dplyr` is to the relational algebra.  The
`plyranges` Bioconductor package implements the language on top of a
key subset of Bioconductor data structures and thus fully integrates
with the Bioconductor framework, gaining access to its scalable data
representations and sophisticated statistical methods.

@@

(ref:GRanges-cap) An illustration of the GRanges data model for a sample from an RNA-seq experiment. The core components of the data model include a seqname column (representing the chromosome), a ranges column which consists of start and end coordinates for a genomic region, and a strand identifier (either positive, negative, or unstranded). Metadata are included as columns to the right of the dotted line as annotations (gene\_id) or range level covariates (score).

@@

The `plyranges` DSL is built on the core Bioconductor data structure
GRanges, which is a constrained table, with fixed columns
for the chromosome, start and end coordinates, and the strand, along
with an arbitrary set of additional columns, consisting of
measurements or metadata specific to the data type or experiment
(figure \@ref(fig:GRanges)).  GRanges balances flexibility with formal
constraints, so that it is applicable to virtually any genomic
workflow, while also being semantically rich enough to support
high-level operations on genomic ranges. As a core data structure,
GRanges enables interoperability between `plyranges` and the rest of
Bioconductor. Adhering to a single data structure simplifies the API
and makes it easier to learn and understand, in part because
operations become endomorphic, i.e., they return the same type as
their input.

GRanges follow the intuitive tidy data pattern: it is a rectangular
table corresponding to a single biological context. Each row contains
a single observation and each column is a variable describing the
observations.  GRanges specializes the tidy pattern in that the
observations always pertain to some genomic feature, but it largely
remains compatible with the general relational operations defined by
`dplyr`. Thus, we define our algebra as an extension of the `dplyr`
algebra, and borrow its syntax conventions and design principles.

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|l|p{6cm}|}
  \hline
  & Verb &  Description \\ 
  \hline
   & \textbf{\emph{summarize()}} & aggregate over column(s) \\ 
   Aggregate & \emph{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
   &  \emph{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
   \hline
   &  \textbf{\emph{mutate()}} & modifies any column \\
   & \textbf{\emph{select()}} & select columns \\
  Modify (Unary) & \textbf{\emph{arrange()}} & sort by columns \\
   & \emph{stretch()} & extend range by fixed amount \\
   &  \emph{shift\_(direction)} & shift coordinates \\
   & \emph{flank\_(direction)} & generate flanking regions \\
   & \emph{\%intersection\% } & row-wise intersection \\
   & \emph{\%union\%} & row-wise union \\
   & \emph{compute\_coverage} & coverage over all ranges \\
  Modify (Binary) &  \emph{\%setdiff\%} & row-wise set difference \\
   & \emph{between()} & row-wise gap range \\
   & \emph{span()} & row-wise spanning range \\
   \hline
    & \emph{join\_overlap\_*()} & merge by overlapping ranges \\
    & \emph{join\_nearest} & merge by nearest neighbor ranges \\
    & \emph{join\_follow} & merge by following ranges \\
    Merge & \emph{join\_precedes} & merge by preceding ranges \\
    & \emph{union\_ranges} & range-wise union \\
    & \emph{intersect\_ranges} & range-wise intersect \\
    & \emph{setdiff\_ranges} & range-wise set difference \\
    & \emph{complement\_ranges} & range-wise set complement \\
  \hline
   & \emph{anchor\_direction()} & fix coordinates at direction \\
  Operate & \textbf{\emph{group\_by()}} & partition by column(s)  \\ 
   & \emph{group\_by\_overlaps()} & partition by overlaps \\
   \hline
   & \textbf{\emph{filter()}} & subset rows \\
  Restrict & \emph{filter\_by\_overlaps()} & subset by overlap \\
    & \emph{filter\_by\_non\_overlaps()} & subset by no overlap \\
   \hline
\end{tabular}
\caption{Overview of the \texttt{plyranges} grammar. The core verbs are
briefly described and categorized into one of the following higher level 
categories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
its origin is from the \texttt{dplyr} grammar.}\label{tab:grammar}
\end{table}

@@

The `plyranges` DSL defines an expressive algebra for performing
genomic operations with and between GRanges objects (see table
\ref{tab:grammar}). The grammar includes several classes of operation
that cover most use cases in genomics data analysis. There are
range arithmetic operators, such as for resizing ranges or finding their
intersection, and operators for merging, filtering and aggregating by
range-specific notions like overlap and proximity.

@@

The algebra also defines conveniences for relative coordinate
adjustments: _shift_ (unanchored adjustment to both start and end) and
_stretch_ (anchored adjustment of width). We can perform any relative
adjustment by some combination of those two operations.  The _stretch_
operation requires an anchor and assumes the midpoint by
default. Since _shift_ is unanchored, the user specifies a suffix for
indicating the direction: left/right or, for stranded features,
upstream/downstream. For example, _shift\_right_ shifts a range to the
right.

@@

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including _intersect_, _union_,
_difference_, and _complement_. There are two sets of these: parallel
and merging. For example, the parallel intersection (_x %intersect% y_) finds the
intersecting range between _xi_ and _yi_ for _i_ in _1...n_, where _n_
is the length of both _x_ and _y_. In contrast, the merging
intersection (_intersect\_ranges(x, y)_) returns a new set of disjoint
ranges representing wherever there was overlap between a range in _x_
and a range in _y_. Finding the parallel union will fail when two ranges
have a gap, so we introduce a _span_ operator that takes the union
while filling any gap. The _complement_ operation is unique in that it
is unary. It finds the regions not covered by any of the ranges in a
single set. Closely related is the _between_ parallel operation, which
finds the gap separating _xi_ and _yi_. The binary operations are
callable from within arithmetic, restriction and aggregation
expressions.

@@

(ref:olaps-cap) Illustration of the three overlap join operators. Each join takes two  GRanges objects, \textit{x} and \textit{y} as input. A 'Hits' object for the  join is computed which consists of two components. The first component contains  the indices of the ranges in \textit{x} that have been overlapped  (the rectangles of \textit{x} that cross the orange lines). The second  component consists of the indices of the ranges in \textit{y} that overlap the ranges in \textit{x}. In this case a range in \textit{y} overlaps the ranges in \textit{x} three times, so the index is repeated three times. The resulting 'Hits' object is used to modify \textit{x} by where it was 'hit' by \textit{y} and merge all metadata columns from \textit{x} and \textit{y} based on the indices contained in the 'Hits' object. This procedure is applied generally in the \texttt{plyranges} DSL for both overlap and nearest neighbor operations. The join semantics alter what is returned: \textbf{A}: for an \textbf{inner} join the \textit{x} ranges that are overlapped by \textit{y} are returned. The returned ranges also include the metadata from the \textit{y} range that overlapped the three \textit{x} ranges. \textbf{B} An \textbf{intersect} join is identical to an inner join except that the intersection is taken between the overlapped \textit{x} ranges and the \textit{y} ranges. \textbf{C} For the \textbf{left} join all \textit{x} ranges are returned regardless of whether they are overlapped by \textit{y}. In this case the third range (rectangle with the asterisk next to it) of the join would have missing values on metadata columns that came from \textit{y}.

To support merging, our algebra recasts finding overlaps or 
nearest neighbors between two
genomic regions as variants of the relational join operator. A join
acts on two GRanges objects: _x_ and _y_. The join operator
is relational in the sense that metadata from the _x_ and _y_
ranges are retained in the joined range.  All join operators in the
`plyranges` DSL generate a set of hits based on overlap or proximity
of ranges and use those hits to merge the two datasets in different
ways. There are four supported matching algorithms: _overlap_,
_nearest_, _precede_, and _follow_ (figure 
\@ref(fig:olaps-fig)). We can further restrict the
matching by whether the query is completely _within_ the subject, and
adding the _directed_ suffix ensures that matching ranges have the
same direction (strand).

@@

(ref:code-comparison-cap) Idiomatic code examples for \texttt{plyranges} (\textbf{A}) and \texttt{GenomicRanges} (\textbf{B}) illustrating an overlap and aggregate operation that returns the same result. In each example, we have two BED files consisting of SNPs that are genome-wide association study (GWAS) hits and reference exons. Each code block counts for each SNP the number of distinct exons it overlaps. The \texttt{plyranges} code achieves this with an overlap join followed by partitioning and aggregation. Strand is ignored by default here. The \texttt{GenomicRanges} code achieves this using the 'Hits' and 'List' classes and their methods.

Since the GRanges object is a tabular data structure, our grammar includes
operators to filter, sort and aggregate by columns in a GRanges. These 
operations can be performed over partitions formed
using the *group_by* modifier. Together with our algebra for arithmetic and
merging, these operations conform to the semantics and syntax of the 
`dplyr` grammar. Consequently, `plyranges` code is generally more compact
than the equivalent `GenomicRanges` code (figure \@ref(fig:code-comparison)).

### Developing workflows with `plyranges`

Here we provide illustrative examples of using the `plyranges` DSL to show
how our grammar could be integrated into genomic data workflows. As we
construct the workflows we show the data output intermittently to assist
the reader in understanding the pipeline steps. The workflows
highlight how interoperability with existing Bioconductor infrastructure, 
enables easy access to public datasets and methods for analysis and 
visualization.

@@

In the workflow of ChIP-seq data analysis, we are interested in finding peaks 
from islands of coverage over chromosome. Here we will use `plyranges`
to call peaks from islands of coverage above 8 then plot the region
surrounding the tallest peak.

Using `plyranges` and the the Bioconductor package `AnnotationHub` [@R-ahub]
we can download and read BigWig files from ChIP-Seq experiments from 
the Human Epigenome Roadmap project [@Roadmap-Epigenomics-Consortium2015-pr]. 
Here we analyse a BigWig file corresponding to H3 lysine 27 trimethylation 
(H3K27Me3) of primary T CD8+ memory cells from peripheral blood, focussing
on coverage islands over chromosome 10.

@@

```{r load-bw, echo = TRUE}
```

@@

```{r chr10-scores-read, echo = TRUE}
```

@@

```{r chr10-scores-reduce, echo = TRUE}
```

Returning to the GRanges object containing normalized coverage scores, 
we filter to find the coordinates of the peak containing the maximum coverage 
score. We can then find a 5000 nt region centered around the maximum position 
by anchoring and modifying the width.

```{r max-score-region}
```

@@

```{r peak_region}
```

(ref:peak-viz-cap) The final result of the `plyranges` operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot.

@@

Another common operation in genomics data analysis is to compute data
summaries over genomic windows. In `plyranges` this can be achieved
via the `group_by_overlaps()` operator.  We bin and count and
find the average GC content of reads from a
H3K27Me3 ChIP-seq experiment by the Human Epigenome Roadmap
Consortium.

@@

Next we only read in alignments that overlap the genomic locations
we are interested in and select the query sequence. Note that
the reading of the BAM file is deferred: only alignments that pass the filter 
are loaded into memory. We can add another column representing the GC proportion 
for each alignment using the `letterFrequency()` function from the `Biostrings`
package [@R-biostrings]. After computing the GC proportion as the score
column, we drop all other columns in the GRanges object.

@@

We have created a GRanges object from genotyping performed on the
H1 cell line, consisting of approximately two million single nucleotide
polymorphisms (SNPs) and short insertion/deletions (indels). The GRanges object consists
of 7 columns, relating to the alleles of a SNP or indel, the B-allele frequency,
log relative intensity of the probes, GC content score over a probe, and the
name of the probe. We can use this information to compute the transition-transversion
ratio, a quality control metric, within each chromosome in GRanges object.

@@

(ref:titv-viz-cap) The final result of computing quality control metrics over the SNP array data with `plyranges`, displayed as a dot plot. Chromosomes are ordered by their estimated transition-transversion ratio. A white reference line is drawn at the expected ratio for a human exome."

@@

The design of `plyranges` adheres to well understood principles of
language and API design: cognitive consistency, cohesion,
endomorphism and expressiveness [@Green1996-qg]. 
To varying degrees, these principles also underlie the
design of `dplyr` and the Bioconductor infrastructure.

We have aimed for `plyranges` to have a simple and direct mapping to the
user's cognitive model, i.e., how the user thinks about the data. This
requires careful selection of the level of abstraction so that the
user can express workflows in the language of genomics. This motivates
the adoption of the tidy GRanges object as our central data
structure. The basic data.frame and `dplyr` tibble lack any notion of
genomic ranges and so could not easily support our genomic grammar,
with its specific verbs for range-oriented data manipulation. Another
example of cognitive consistency is how `plyranges` is insensitive to
direction/strand by default when, e.g., detecting
overlaps. `GenomicRanges` has the opposite behavior. We believe that
defaulting to purely spatial overlap is most intuitive to most users.

To further enable cognitive consistency, `plyranges` functions are cohesive. 
A function is defined to be cohesive if it performs a singular task without
producing any side-effects. Singular tasks can always be broken 
down further at lower levels of abstraction.
For example, to resize a range, the user needs to specify which
position (start, end, midpoint) should be invariant over the
transformation. The `resize()` function from the `GenomicRanges` package
has a `fix` argument that sets the anchor, so calling `resize()`
coalesces anchoring and width modification. The coupling at the
function call level is justified since the effect of setting the width
depends on the anchor. However, `plyranges` increases cohesion and
decouples the anchoring into its own function call.

Increasing cohesion simplifies the interface to each operation, makes
the meaning of arguments more intuitive, and relies on function names
as the primary means of expression, instead of a more complex mixture
of function and argument names. This results in the
user being able to conceptualize the `plyranges` DSL as a flat catalog of 
functions, without having to descend further into documentation to understand
a function's arguments. A flat function catalog also enhances API
discoverability, particularly through auto-completion in 
integrated developer environments (IDEs). One downside of pushing cohesion to 
this extreme is that function calls become coupled, and care is 
necessary to treat them as a group when modifying code.

Like `dplyr`, `plyranges` verbs are functional: they are free of side
effects and are generally endomorphic, meaning that when the input is a GRanges
object they return a GRanges object. This enables chaining of verbs
through syntax like the forward pipe operator from the `magrittr` package. 
This syntax has a direct cognitive mapping to natural language and the intuitive notion of
pipelines. The low-level object-oriented APIs of Bioconductor tend to
manipulate data via sub-replacement functions, like `start(gr) <-
x`. These ultimately produce the side effect of replacing a symbol
mapping in the current environment and thus are not amenable to so-called 
fluent syntax.

>>>>

There is a cognitive load placed on users in learning a data abstraction from the Bioconductor project and understanding its appropriate use. Users must navigate these abstractions to perform a genomic analysis task, when a single data abstraction, a _GRanges_ object will suffice. By recognizing that the _GRanges_ class follows 'tidy' data principles, we create a grammar of genomic data transformation, defining verbs for performing actions on and between genomic interval data and providing a way of performing common data analysis tasks through a coherent interface to existing Bioconductor infrastructure. We implement this grammar as a Bioconductor/R package called **plyranges**.

@@

Several domain specific languages (DSLs) enable scientists to process and reason
about heterogeneous genomics data by expressing common
operations, such as range manipulation and overlap-based joins, using
the vocabulary of genomics. Their implementations either delegate
computations to a database, or operate over collections of files in
standard formats like BED. An example of the former is the 
**Genome Query Language (GQL)** and its distributed implementation **GenAp** 
which use a  SQL-like syntax for fast retrieval of information of unprocessed 
sequencing data [@Kozanitis2014-va; @Kozanitis2016-bm]. Similarly, the 
**Genometric Query Language (GMQL)** implements a DSL for combining genomic 
datasets [@Kaitoua2017-pw].  The command line application **BEDtools** develops 
an extensive algebra for performing arithmetic between two or more sets of 
genomic regions [@Quinlan2010-gc]. All of the aforementioned DSLs are 
designed to be evaluated either at the command line or embedded in scripts 
for batch processing. They exist in a sparse ecosystem, mostly consisting of 
UNIX and database tools that lack biological semantics and operate at the 
level of files and database tables.

The Bioconductor/R packages **IRanges** and **GenomicRanges** 
[@r-core; @Lawrence2013-wg; @Huber2015-ei] define a DSL for 
analyzing genomics data with R, an interactive data analysis environment that 
encourages reproducibility and provides high-level abstractions for manipulating,
modelling and plotting data, through state of the art methods in
statistical computing. The packages define object-oriented (OO)
abstractions for representing genomic data and enable
interoperability by allowing users and developers to use these
abstractions in their own code and packages. Other genomic DSLs 
that are embedded in programming languages include **pybedtools** and **valr** 
[@Dale2011-js; @Kent2017], however these packages lack the interoperability
provided by the aforementioned Bioconductor packages and are not easily 
extended.

@@

Recently, there has been a concerted, community effort to standardize
R data structures and workflows around the notion of tidy data
[@Wickham2014-jc]. A tidy dataset is defined as a tabular
data structure that has observations as rows and columns as variables,
and all measurements pertain to a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relate to the design of an experiment
and the variables measured. The **dplyr** package [@Wickham2017-dplyr] 
defines an application programming interface (API) that maps notions from 
the general relational algebra to verbs that act on tidy data. These verbs
can be composed together on one or more tidy datasets with the pipe operator 
from the **magrittr** package [@R-magrittr]. Taken together these features 
enable a user to write human readable analysis workflows.

We have created a genomic DSL called **plyranges** that reformulates
notions from existing genomic algebras and embeds them in R as a
genomic extension of **dplyr**. By analogy, **plyranges** is to the
genomic algebra, as **dply**` is to the relational algebra.  The
**plyranges`**Bioconductor package implements the language on top of a
key subset of Bioconductor data structures and thus fully integrates
with the Bioconductor framework, gaining access to its scalable data
representations and sophisticated statistical methods.

@@

(ref:GRanges-cap) An illustration of the _GRanges_ data model for a sample from an RNA-seq experiment. The core components of the data model include a seqname column (representing the chromosome), a ranges column which consists of start and end coordinates for a genomic region, and a strand identifier (either positive, negative, or unstranded). Metadata are included as columns to the right of the dotted line as annotations (gene\_id) or range level covariates (score).

@@

The **plyranges** DSL is built on the core Bioconductor data structure
_GRanges_, which is a constrained table, with fixed columns
for the chromosome, start and end coordinates, and the strand, along
with an arbitrary set of additional columns, consisting of
measurements or metadata specific to the data type or experiment
(figure \@ref(fig:GRanges)).  _GRanges_ balances flexibility with formal
constraints, so that it is applicable to virtually any genomic
workflow, while also being semantically rich enough to support
high-level operations on genomic ranges. As a core data structure,
_GRanges_ enables interoperability between **plyranges** and the rest of
Bioconductor. Adhering to a single data structure simplifies the API
and makes it easier to learn and understand, in part because
operations become endomorphic, i.e., they return the same type as
their input.

_GRanges_ follow the intuitive tidy data pattern: it is a rectangular
table corresponding to a single biological context. Each row contains
a single observation and each column is a variable describing the
observations.  _GRanges_ specializes the tidy pattern in that the
observations always pertain to some genomic feature, but it largely
remains compatible with the general relational operations defined by
**dplyr**. Thus, we define our algebra as an extension of the **dplyr**
algebra, and borrow its syntax conventions and design principles.

\begin{table}[!htbp]
\centering
\begin{tabular}{|l|l|p{6cm}|}
  \hline
  & Verb &  Description \\ 
  \hline
   & \textbf{\emph{summarize()}} & aggregate over column(s) \\ 
   Aggregate & \textt{disjoin\_ranges()} & aggregate column(s) over the union of end coordinates \\
   &  \textt{reduce\_ranges()} & aggregate column(s) by merging overlapping and neighboring ranges \\
   \hline
   &  \textbf{\textt{mutate()}} & modifies any column \\
   & \textbf{\textt{select()}} & select columns \\
  Modify (Unary) & \textbf{\textt{arrange()}} & sort by columns \\
   & \textt{stretch()} & extend range by fixed amount \\
   &  \textt{shift\_(direction)} & shift coordinates \\
   & \textt{flank\_(direction)} & generate flanking regions \\
   & \textt{\%intersection\% } & row-wise intersection \\
   & \textt{\%union\%} & row-wise union \\
   & \textt{compute\_coverage} & coverage over all ranges \\
  Modify (Binary) &  \textt{\%setdiff\%} & row-wise set difference \\
   & \textt{between()} & row-wise gap range \\
   & \textt{span()} & row-wise spanning range \\
   \hline
    & \textt{join\_overlap\_*()} & merge by overlapping ranges \\
    & \textt{join\_nearest} & merge by nearest neighbor ranges \\
    & \textt{join\_follow} & merge by following ranges \\
    Merge & \textt{join\_precedes} & merge by preceding ranges \\
    & \textt{union\_ranges} & range-wise union \\
    & \textt{intersect\_ranges} & range-wise intersect \\
    & \textt{setdiff\_ranges} & range-wise set difference \\
    & \textt{complement\_ranges} & range-wise set complement \\
  \hline
   & \textt{anchor\_direction()} & fix coordinates at direction \\
  Operate & \textbf{\textt{group\_by()}} & partition by column(s)  \\ 
   & \textt{group\_by\_overlaps()} & partition by overlaps \\
   \hline
   & \textbf{\textt{filter()}} & subset rows \\
  Restrict & \textt{filter\_by\_overlaps()} & subset by overlap \\
    & \textt{filter\_by\_non\_overlaps()} & subset by no overlap \\
   \hline
\end{tabular}
\caption{Overview of the \textbf{plyranges} grammar. The core verbs are
briefly described and categorized into one of the following higher level 
categories: aggregate, modify, merge, operate, or restrict. A verb is given bold text if
its origin is from the \textbf{dplyr} grammar.}\label{tab:grammar}
\end{table}

@@

The **plyranges** DSL defines an expressive algebra for performing
genomic operations with and between _GRanges_ objects (see table
\ref{tab:grammar}). The grammar includes several classes of operation
that cover most use cases in genomics data analysis. There are
range arithmetic operators, such as for resizing ranges or finding their
intersection, and operators for merging, filtering and aggregating by
range-specific notions like overlap and proximity.

@@

The algebra also defines conveniences for relative coordinate
adjustments: _shift_ (unanchored adjustment to both start and end) and
_stretch_ (anchored adjustment of width). We can perform any relative
adjustment by some combination of those two operations.  The _stretch_
operation requires an anchor and assumes the midpoint by
default. Since _shift_ is unanchored, the user specifies a suffix for
indicating the direction: left/right or, for stranded features,
upstream/downstream. For example, `shift_right()` shifts a range to the
right.

@@

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including _intersect_, _union_,
_difference_, and _complement_. There are two sets of these: parallel
and merging. For example, the parallel intersection (`x %intersect% y`) finds the
intersecting range between _xi_ and _yi_ for _i_ in _1...n_, where _n_
is the length of both _x_ and _y_. In contrast, the merging
intersection ( `intersect_ranges(x, y)`) returns a new set of disjoint
ranges representing wherever there was overlap between a range in _x_
and a range in _y_. Finding the parallel union will fail when two ranges
have a gap, so we introduce a `span()` operator that takes the union
while filling any gap. The `complement()` operation is unique in that it
is unary. It finds the regions not covered by any of the ranges in a
single set. Closely related is the `between()` parallel operation, which
finds the gap separating _xi_ and _yi_. The binary operations are
callable from within arithmetic, restriction and aggregation
expressions.

@@

(ref:olaps-cap) Illustration of the three overlap join operators. Each join takes two  _GRanges_ objects, \textit{x} and \textit{y} as input. A 'Hits' object for the  join is computed which consists of two components. The first component contains  the indices of the ranges in \textit{x} that have been overlapped  (the rectangles of \textit{x} that cross the orange lines). The second  component consists of the indices of the ranges in \textit{y} that overlap the ranges in \textit{x}. In this case a range in \textit{y} overlaps the ranges in \textit{x} three times, so the index is repeated three times. The resulting 'Hits' object is used to modify \textit{x} by where it was 'hit' by \textit{y} and merge all metadata columns from \textit{x} and \textit{y} based on the indices contained in the 'Hits' object. This procedure is applied generally in the \textbf{plyranges} DSL for both overlap and nearest neighbor operations. The join semantics alter what is returned: \textbf{A}: for an \textbf{inner} join the \textit{x} ranges that are overlapped by \textit{y} are returned. The returned ranges also include the metadata from the \textit{y} range that overlapped the three \textit{x} ranges. \textbf{B} An \textbf{intersect} join is identical to an inner join except that the intersection is taken between the overlapped \textit{x} ranges and the \textit{y} ranges. \textbf{C} For the \textbf{left} join all \textit{x} ranges are returned regardless of whether they are overlapped by \textit{y}. In this case the third range (rectangle with the asterisk next to it) of the join would have missing values on metadata columns that came from \textit{y}.

To support merging, our algebra recasts finding overlaps or 
nearest neighbors between two
genomic regions as variants of the relational join operator. A join
acts on two _GRanges_ objects: _x_ and _y_. The join operator
is relational in the sense that metadata from the _x_ and _y_
ranges are retained in the joined range.  All join operators in the
`plyranges` DSL generate a set of hits based on overlap or proximity
of ranges and use those hits to merge the two datasets in different
ways. There are four supported matching algorithms: _overlap_,
_nearest_, _precede_, and _follow_ (figure 
\@ref(fig:olaps-fig)). We can further restrict the
matching by whether the query is completely _within_ the subject, and
adding the _directed_ suffix ensures that matching ranges have the
same direction (strand).

@@

(ref:code-comparison-cap) Idiomatic code examples for \textbf{plyranges} (\textbf{A}) and \textbf{GenomicRanges} (\textbf{B}) illustrating an overlap and aggregate operation that returns the same result. In each example, we have two BED files consisting of SNPs that are genome-wide association study (GWAS) hits and reference exons. Each code block counts for each SNP the number of distinct exons it overlaps. The \textbf{plyranges} code achieves this with an overlap join followed by partitioning and aggregation. Strand is ignored by default here. The \textbf{GenomicRanges} code achieves this using the _Hits_ and _List_ classes and their methods.

Since the _GRanges_ object is a tabular data structure, our grammar includes
operators to filter, sort and aggregate by columns in a _GRanges_. These 
operations can be performed over partitions formed
using the `group_by()` modifier. Together with our algebra for arithmetic and
merging, these operations conform to the semantics and syntax of the 
**dplyr** grammar. Consequently, **plyranges** code is generally more compact
than the equivalent **GenomicRanges** code (figure \@ref(fig:code-comparison)).

### Developing workflows with **plyranges**

Here we provide illustrative examples of using the **plyranges** DSL to show
how our grammar could be integrated into genomic data workflows. As we
construct the workflows we show the data output intermittently to assist
the reader in understanding the pipeline steps. The workflows
highlight how interoperability with existing Bioconductor infrastructure, 
enables easy access to public datasets and methods for analysis and 
visualization.

@@

In the workflow of ChIP-seq data analysis, we are interested in finding peaks 
from islands of coverage over chromosome. Here we will use **plyranges**
to call peaks from islands of coverage above 8 then plot the region
surrounding the tallest peak.

Using **plyranges** and the the Bioconductor package **AnnotationHub** [@R-ahub]
we can download and read BigWig files from ChIP-Seq experiments from 
the Human Epigenome Roadmap project [@Roadmap-Epigenomics-Consortium2015-pr]. 
Here we analyse a BigWig file corresponding to H3 lysine 27 trimethylation 
(H3K27Me3) of primary T CD8+ memory cells from peripheral blood, focussing
on coverage islands over chromosome 10.

@@

```{r load-bw, echo = TRUE }
```

@@

```{r chr10-scores-read, echo = TRUE }
```

@@

```{r chr10-scores-reduce, echo = TRUE }
```

Returning to the _GRanges_ object containing normalized coverage scores, 
we filter to find the coordinates of the peak containing the maximum coverage 
score. We can then find a 5000 nt region centered around the maximum position 
by anchoring and modifying the width.

```{r max-score-region }
```

@@

```{r peak_region }
```

(ref:peak-viz-cap) The final result of the **plyranges** operations to find a 5000nt region surrounding the peak of normalised coverage scores over chromosome 10, displayed as a density plot.

@@

Another common operation in genomics data analysis is to compute data
summaries over genomic windows. In **plyranges** this can be achieved
via the `group_by_overlaps()` operator.  We bin and count and
find the average GC content of reads from a
H3K27Me3 ChIP-seq experiment by the Human Epigenome Roadmap
Consortium.

@@

Next we only read in alignments that overlap the genomic locations
we are interested in and select the query sequence. Note that
the reading of the BAM file is deferred: only alignments that pass the filter 
are loaded into memory. We can add another column representing the GC proportion 
for each alignment using the `letterFrequency()` function from the **Biostrings**
package [@R-biostrings]. After computing the GC proportion as the score
column, we drop all other columns in the _GRanges_ object.

@@

We have created a _GRanges_ object from genotyping performed on the
H1 cell line, consisting of approximately two million single nucleotide
polymorphisms (SNPs) and short insertion/deletions (indels). The _GRanges_ object consists
of 7 columns, relating to the alleles of a SNP or indel, the B-allele frequency,
log relative intensity of the probes, GC content score over a probe, and the
name of the probe. We can use this information to compute the transition-transversion
ratio, a quality control metric, within each chromosome in _GRanges_ object.

@@

(ref:titv-viz-cap) The final result of computing quality control metrics over the SNP array data with **plyranges**, displayed as a dot plot. Chromosomes are ordered by their estimated transition-transversion ratio. A white reference line is drawn at the expected ratio for a human exome."

@@

The design of **plyranges** adheres to well understood principles of
language and API design: cognitive consistency, cohesion,
endomorphism and expressiveness [@Green1996-qg]. 
To varying degrees, these principles also underlie the
design of **dplyr** and the Bioconductor infrastructure.

We have aimed for **plyranges** to have a simple and direct mapping to the
user's cognitive model, i.e., how the user thinks about the data. This
requires careful selection of the level of abstraction so that the
user can express workflows in the language of genomics. This motivates
the adoption of the tidy _GRanges_ object as our central data
structure. The basic _data.frame_ and **dplyr** _tibble_ lack any notion of
genomic ranges and so could not easily support our genomic grammar,
with its specific verbs for range-oriented data manipulation. Another
example of cognitive consistency is how **plyranges** is insensitive to
direction/strand by default when, e.g., detecting
overlaps. **GenomicRanges** has the opposite behavior. We believe that
defaulting to purely spatial overlap is most intuitive to most users.

To further enable cognitive consistency, **plyranges** functions are cohesive. 
A function is defined to be cohesive if it performs a singular task without
producing any side-effects. Singular tasks can always be broken 
down further at lower levels of abstraction.
For example, to resize a range, the user needs to specify which
position (start, end, midpoint) should be invariant over the
transformation. The `resize()` function from the **GenomicRanges** package
has a `fix` argument that sets the anchor, so calling `resize()`
coalesces anchoring and width modification. The coupling at the
function call level is justified since the effect of setting the width
depends on the anchor. However, **plyranges** increases cohesion and
decouples the anchoring into its own function call.

Increasing cohesion simplifies the interface to each operation, makes
the meaning of arguments more intuitive, and relies on function names
as the primary means of expression, instead of a more complex mixture
of function and argument names. This results in the
user being able to conceptualize the **plyranges** DSL as a flat catalog of 
functions, without having to descend further into documentation to understand
a function's arguments. A flat function catalog also enhances API
discoverability, particularly through auto-completion in 
integrated developer environments (IDEs). One downside of pushing cohesion to 
this extreme is that function calls become coupled, and care is 
necessary to treat them as a group when modifying code.

Like **dplyr**, **plyranges** verbs are functional: they are free of side
effects and are generally endomorphic, meaning that when the input is a _GRanges_
object they return a _GRanges_ object. This enables chaining of verbs
through syntax like the forward pipe operator from the **magrittr** package. 
This syntax has a direct cognitive mapping to natural language and the intuitive notion of
pipelines. The low-level object-oriented APIs of Bioconductor tend to
manipulate data via sub-replacement functions, like `start(gr) <-
x`. These ultimately produce the side effect of replacing a symbol
mapping in the current environment and thus are not amenable to so-called 
fluent syntax.
